---
title: "ggradar Tidy Tuesday Demo"
author: "Cameryn Brock"
date: "5/25/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## `ggradar`

https://github.com/ricardo-bion/ggradar

If you need to download, can download with 
`devtools::install_github("ricardo-bion/ggradar", dependencies = TRUE)`
  

We will also be using `calecopal`

https://github.com/an-bui/calecopal

If you need to download, can download with 
`devtools::install_github("an-bui/calecopal")`

See all palette names with `names(cal_palettes)`


```{r}
library(palmerpenguins)
library(ggradar)
library(calecopal)
library(tidyverse)
library(janitor)
library(tidytuesdayR)
```

### Read in data with `tidytuesdayR`

Information for tidy tuesday data sets available here: https://github.com/rfordatascience/tidytuesday

We will be using data for global crop yields from Our World in Data. https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-01/readme.md


```{r}
crops <- tidytuesdayR::tt_load('2020-09-01')

crops_clean <- crops$key_crop_yields %>% 
  clean_names() %>% 
  rename_with(~str_remove(., pattern = "_tonnes_per_hectare"))
```


```{r}
# warning that 'entity' contains countries, continents, and regions
# unique(crops$entity)
# filter to just continent-level
# select only crops of interest

crops_focal <- crops_clean %>%
  filter(
    entity %in% c("Northern America",
                  "Central America",
                  "South America",
                  "Europe",
                  "Asia",
                  "Europe",
                  "Africa")) %>%
  rename("region" = entity) %>%
  mutate(region = factor(
    region,
    levels = c(
      "Northern America",
      "Central America",
      "South America",
      "Europe",
      "Asia",
      "Africa"))) %>% 
select(region, 
         year,
         wheat,
         rice,
         maize,
         soybeans,
         potatoes,
         beans,
         bananas)
```

```{r}
# store max values for later

max_wheat <- max(crops_focal$wheat, na.rm = TRUE)
max_rice <- max(crops_focal$rice, na.rm = TRUE)
max_maize <- max(crops_focal$maize, na.rm = TRUE)
max_soy <- max(crops_focal$soybeans, na.rm = TRUE)
max_pot <- max(crops_focal$potatoes, na.rm = TRUE)
max_beans <- max(crops_focal$beans, na.rm = TRUE)
max_ban <- max(crops_focal$bananas, na.rm = TRUE)
```

```{r}
# summarize by mean value

crops_2018 <- crops_focal %>% 
  filter(year == 2018) %>% 
  select(!year)
```

```{r}
# standardize by max value

crops_stand_2018 <- crops_2018 %>% 
  mutate(wheat = wheat/max_wheat,
         rice = rice/max_rice,
         maize = maize/max_maize,
         soybeans = soybeans/max_soy,
         potatoes = potatoes/max_pot,
         beans = beans/max_beans,
         bananas = bananas/max_ban)
```

## Visualize!

You may be thinking you can just make a line graph in ggplot and use coord_polar...

```{r}
crops_ggplot <- crops_stand_2018 %>% 
  pivot_longer(2:8,
               names_to = "crop",
               values_to = "stand")

ggplot(crops_ggplot, aes(x = crop, y = stand, group = region, color = region)) + 
  geom_line()
  
ggplot(crops_ggplot, aes(x = crop, y = stand, group = region, color = region)) + 
  geom_line() + 
  coord_polar()
```



```{r}
# generic output
ggradar(crops_stand)
```

```{r}
# Visualize!
# colors that represent region
my_cat_colors <- cal_palette("figmtn")
my_seq_colors <- cal_palette("sierra1", n = 11, type = "continuous")

# pretty labels
my_labels = c("Wheat", "Rice", "Maize", 
              "Soybeans", "Potatoes", "Beans",
              "Bananas")
```


```{r, fig.width = 10, fig.height = 6}
ggradar(crops_stand,
        axis.labels = my_labels,
        background.circle.transparency = 0.05,
        gridline.min.colour = "gray70",
        gridline.mid.colour = "gray60",
        gridline.max.colour = "gray50") + 
  scale_color_manual(values = my_cat_colors)
```

Another way to look at this is by year per country...

```{r}
# Central America

crops_ca <- crops_focal %>%
  filter(region == "Central America")

crops_ca_stand <- crops_ca %>% 
  select(!region) %>%
  mutate(
    wheat = wheat / max_wheat,
    rice = rice / max_rice,
    maize = maize / max_maize,
    soybeans = soybeans / max_soy,
    potatoes = potatoes / max_pot,
    beans = beans / max_beans,
    bananas = bananas / max_ban) %>% 
  filter(year %in% c(seq(from = 1960, to = 2020, by = 5)))

ggradar(crops_ca_stand,
        axis.labels = my_labels,
        background.circle.transparency = 0.05,
        gridline.min.colour = "gray70",
        gridline.mid.colour = "gray60",
        gridline.max.colour = "gray50") + 
  scale_color_manual(values = my_seq_colors)
```

```{r}
## North America

crops_na <- crops_focal %>%
  filter(region == "Northern America") %>% 
  select(!region) %>%
  mutate(
    wheat = wheat / max_wheat,
    rice = rice / max_rice,
    maize = maize / max_maize,
    soybeans = soybeans / max_soy,
    potatoes = potatoes / max_pot,
    beans = beans / max_beans,
    bananas = bananas / max_ban) %>% 
  filter(year %in% c(seq(from = 1960, to = 2020, by = 5)))

ggradar(crops_na,
        axis.labels = my_labels,
        background.circle.transparency = 0.05,
        gridline.min.colour = "gray70",
        gridline.mid.colour = "gray60",
        gridline.max.colour = "gray50") + 
  scale_color_manual(values = my_seq_colors)
```

```{r}
# Africa

crops_af <- crops_focal %>%
  filter(region == "Africa") %>% 
  select(!region) %>%
  mutate(
    wheat = wheat / max_wheat,
    rice = rice / max_rice,
    maize = maize / max_maize,
    soybeans = soybeans / max_soy,
    potatoes = potatoes / max_pot,
    beans = beans / max_beans,
    bananas = bananas / max_ban) %>% 
  filter(year %in% c(seq(from = 1960, to = 2020, by = 5)))

ggradar(crops_af,
        axis.labels = my_labels,
        background.circle.transparency = 0.05,
        gridline.min.colour = "gray70",
        gridline.mid.colour = "gray60",
        gridline.max.colour = "gray50") + 
  scale_color_manual(values = my_seq_colors) 
```



## Caveats to using radar charts

https://www.data-to-viz.com/caveat/spider.html

* Circular layout is hard to read
* Difficult to find ranking with no clear order
* Category order has large impact
* Scales can be confusing
* Difficult to evaluate differences
* Overplotting can make chart hard to read

## Good application example: 

Clark, M. A., Springmann, M., Hill, J., & Tilman, D. (2019). Multiple health and environmental impacts of foods. Proceedings of the National Academy of Sciences, 116(46), 23357–23362. https://doi.org/10.1073/pnas.1906908116